Analysis of Programming Students' Prompts to Identify Patterns of Critical Thinking in Interactions with Generative AI \\ {\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and should not be used} \thanks{Identify applicable funding agency here. If none, delete this.}

### 2024-08-05 paper/paper.tex
first draft

\title{An In-Depth Analysis of Student Prompts With CharlieBot \\
{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
should not be used}
\thanks{Identify applicable funding agency here. If none, delete this.}
}

@@

According \cite{LO2024105100} further research about the use of AI tools in
education is needed to understand how these tools can be used to improve
students' learning experience. In particular, the study suggested that future
research on students' study habits.

@@

The paper \cite{10343037} examined students from different demographic groups
using AI tools like ChatGPT, comparing perceptions and impacts
among these groups. The results showed that AI tools had a positive impact on
academic performance, especially for students with less prior experience.

@@

The table \ref{tab:categories} shows the categories of analysis used in this
study. The categories were based on the work of
\cite{10.1007/978-3-031-64299-9_20}, with the addition of four new categories.
Also, we group the categories to understand the impact on critical thinking.

@@

CharlieBot is a ChatGPT-based bot that provides assistance to students in
Java programming courses in UC3M. CharlieBot logs all conversations with
students, which were used for this study. The goal was to categorize student
messages to understand how students interact with the bot and identify patterns
in their study habits.
\\
The categories outlined in \cite{10.1007/978-3-031-64299-9_20} served as the
foundation for classifying student messages. However, it was identified that
some categories were missing, necessitating the addition of four new categories.
Table \ref{tab:categories} presents these categories along with their
descriptions.
\\
Subsequently, these eight categories were input into Claude.ai, accompanied by
detailed instructions on the desired structure of the input and the expected
format of the output. For each conversation with the CharlieBot log, a .csv file
was downloaded, and Claude was tasked with classifying the user messages into
one of the eight categories.

@@

Approximately 64\% of the messages pertain to categories associated with
critical thinking, corroborating with the findings of
\cite{10.1007/978-3-031-64299-9_20}. In contrast, around 28\% of the messages
indicate a preference among students for ready-made answers.

>>>>

\title{Analysis of Programming Students' Prompts to Identify Patterns of
Critical Thinking in Interactions with Generative AI \\
{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
should not be used}
\thanks{Identify applicable funding agency here. If none, delete this.}
}

@@

According \cite{Kwan24} further research about the use of AI tools in
education is needed to understand how these tools can be used to improve
students' learning experience. In particular, the study suggested that future
research on students' study habits.

Problems

The literature suggests a concern regarding the impact of ChatGPT on students'
critical thinking abilities. Vargas \cite{Vargas2023} indicated that
ChatGPT use might lead to an overreliance on the tool, potentially diminishing
students' critical thinking skills. Similarly, Cai \cite{cai23} identified
overdependence and reduced intellectual engagement as significant drawbacks of
ChatGPT-supported learning. Chan \cite{chan23} also noted that this
reliance could result in a decrease in critical thinking, as students might make
decisions based solely on the information provided by ChatGPT. Collectively,
these studies highlight a common theme of reduced critical engagement and
overreliance on ChatGPT in educational contexts.

@@

The paper \cite{Kwan24} examined students from different demographic groups
using AI tools like ChatGPT, comparing perceptions and impacts
among these groups. The results showed that AI tools had a positive impact on
academic performance, especially for students with less prior experience.

\subsection{Critical Thinking}

Critical thinking is a fundamental skill for students to develop, as it enables
them to analyze, evaluate, and synthesize information. This skill is essential
for problem-solving and decision-making, making it a crucial aspect of
educational success, specifically in programming courses. However, the use of AI
tools like ChatGPT has raised concerns about the potential impact on students'
critical thinking abilities \cite{Vargas2023} \cite{cai23} \cite{chan23}.

In the other hand, the study of \cite{zhang24} suggested that incorporating
ChatGPT could stimulate students' critical thinking, probably because they
had to assess the correctness of ChatGPT's answers.

Regarding self-regulation, \cite{cai23} noted that ChatGPT allowed students to
regulate their learning pace.

\cite{wu24} found that ChatGPT-supported learning not only promoted students'
self-regulation but also increased their self-efficacy.

Finally, some students asserted that the use of ChatGPT enhanced the depth of
their thinking and understanding \cite{chan23}.

\subsection{CharlieBot}

CharlieBot is a ChatGPT-based bot that provides assistance to students in Java
programming courses at UC3M. The bot uses RAG(retrieval augmented generation) to
generate responses to student queries, providing explanations, code snippets,
and debugging help. The goal of CharlieBot is to support students in their
learning process by providing personalized assistance and feedback.

@@

The table \ref{tab:categories} shows the categories of analysis used in this
study. The categories were based on the work of
\cite{Ghimire24} with the addition of four new categories.
Also, we group the categories to understand the impact on critical thinking.

@@

CharlieBot is a ChatGPT-based bot that provides assistance to students in
Java programming courses in UC3M. CharlieBot logs all conversations with
students, which were used for this study. The goal was to categorize student
messages to understand how students interact with the bot and identify patterns
in their study habits.
\\
The categories outlined in \cite{Ghimire24} served as the
foundation for classifying student messages (Debugging Help, Conceptual
Questions, Code Snippet and Complete Solution). However, it was identified that
some categories were missing, necessitating the addition of new categories
(Student Correction, Multiple Question, Language Change and Uncategorized). Table
\ref{tab:categories} presents these categories along with their descriptions.
\\
Subsequently, these eight categories were input into Claude.ai, accompanied by
detailed instructions on the desired structure of the input and the expected
format of the output. For each conversation with the CharlieBot log, a .csv file
was downloaded, and Claude was tasked with classifying the user messages into
one of the eight categories.

@@

Approximately 64\% of the messages pertain to categories associated with
critical thinking, corroborating with the findings of
\cite{Ghimire24}. In contrast, around 28\% of the messages
indicate a preference among students for ready-made answers.
